---
title: "Student Inflow Analysis - MVP Prototype"
format: html
---

```{r}
#| label: setup
#| cache: false
#| output: false
#| include: false
#| freeze: false

# Load required libraries
source("00_setup.R")

```

## 1. Data Loading & Preparation

First, let's load the application data and reference tables:

```{r}
#| label: load-data

applications <- suppressWarnings(safe_read_csv("data/cambo/aanmeldingen_oktober_2024.csv", 
                      delim = ",",
                      col_types = cols(.default = col_guess())))

# Load reference data
brin <- read_excel("data/reference/BRIN.xlsx")
programmes <- read_excel("data/reference/Opleiding_dimensie.xlsx")

# Join with reference data
applications_enriched <- applications %>%
    # Convert opleidingcode to integer for joining (drops a small % of non-numeric values)
    mutate(opleidingcode = as.integer(opleidingcode)) %>%
    left_join(programmes, by = c("opleidingcode" = "Opleidingcode")) %>%
    left_join(brin, by = c("instellingserkenningscode" = "BRIN"))
```

## 2. Application Overview Dashboard

Let's create key KPIs and visualizations that would form our main dashboard:

```{r}
#| label: dashboard-metrics

# Calculate key metrics by year
yearly_metrics <- applications_enriched %>%
  group_by(schooljaar) %>%
  summarise(
    total_applications = n(),
    unique_students = n_distinct(bsnhash),
    conversion_rate = mean(status == "ENROLLED", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  arrange(schooljaar)

# Display metrics table
knitr::kable(yearly_metrics, 
             col.names = c("School Year", "Total Applications", "Unique Students", "Conversion Rate (%)"),
             digits = c(0, 0, 0, 1))
```

```{r}
#| label: yearly-trend-plot

# Create trend visualization
yearly_metrics %>%
  pivot_longer(cols = c(total_applications, unique_students), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(metric = if_else(metric == "total_applications", 
                         "Total Applications", 
                         "Unique Students")) %>%
  ggplot(aes(x = factor(schooljaar), y = value, group = metric, color = metric)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Application Trends by Year",
       x = "School Year",
       y = "Count",
       color = "Metric") +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "bottom")
```

```{r}
#| label: conversion-rate-plot

# Conversion rate visualization
ggplot(yearly_metrics, aes(x = factor(schooljaar), y = conversion_rate)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(conversion_rate, 1), "%")), 
            vjust = -0.5, size = 3.5) +
  theme_minimal() +
  labs(title = "Conversion Rate by Year",
       x = "School Year",
       y = "Conversion Rate (%)") +
  ylim(0, max(yearly_metrics$conversion_rate) * 1.1)
```

## 3. Double Application Analysis

One of our key features is analyzing students who apply to multiple programs or institutions:

```{r}
#| label: double-applications

# Identify and analyze double applications
application_counts <- applications_enriched %>%
  group_by(bsnhash, schooljaar) %>%
  summarise(
    application_count = n(),
    institutions_count = n_distinct(instellingserkenningscode),
    programs_count = n_distinct(opleidingcode),
    is_enrolled = any(status == "ENROLLED"),
    .groups = "drop"
  )

# Distribution of applications per student
app_distribution <- application_counts %>%
  group_by(application_count) %>%
  summarise(
    students = n(),
    converted = sum(is_enrolled),
    conversion_rate = mean(is_enrolled) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    students_pct = students / sum(students) * 100
  )

# Display the distribution
knitr::kable(app_distribution, 
             col.names = c("Applications per Student", "Student Count", 
                          "Enrolled", "Conversion Rate (%)", "% of Total"),
             digits = c(0, 0, 0, 1, 1))
```

```{r}
#| label: app-count-distribution

# Visualization: Application count distribution
ggplot(app_distribution, aes(x = factor(application_count), y = students)) +
  geom_col(fill = "darkcyan") +
  geom_text(aes(label = paste0(round(students_pct, 1), "%")), 
            vjust = -0.5, size = 3) +
  theme_minimal() +
  labs(title = "Distribution of Applications per Student",
       x = "Number of Applications",
       y = "Number of Students") +
  scale_y_continuous(labels = comma)
```

```{r}
#| label: conversion-by-app-count

# Conversion rates by application count
ggplot(app_distribution, aes(x = factor(application_count), y = conversion_rate)) +
  geom_col(fill = "darkgoldenrod") +
  geom_text(aes(label = paste0(round(conversion_rate, 1), "%")), 
            vjust = -0.5, size = 3) +
  theme_minimal() +
  labs(title = "Conversion Rate by Number of Applications",
       x = "Number of Applications",
       y = "Conversion Rate (%)") +
  ylim(0, max(app_distribution$conversion_rate) * 1.1)
```

## 4. Program and Institution Analysis

Let's analyze which programs and institutions have the highest application volumes and conversion rates:

```{r}
#| label: program-analysis

# Top programs by application volume
top_programs <- applications_enriched %>%
  filter(!is.na(Opleidingsnaam)) %>%
  group_by(opleidingcode, Opleidingsnaam) %>%
  summarise(
    applications = n(),
    conversions = sum(status == "ENROLLED", na.rm = TRUE),
    conversion_rate = mean(status == "ENROLLED", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(applications)) %>%
  head(10)

# Display top programs
knitr::kable(top_programs, 
             col.names = c("Code", "Program Name", "Applications", 
                          "Enrollments", "Conversion Rate (%)"),
             digits = c(0, 0, 0, 0, 1))
```

```{r}
#| label: institution-analysis

# Top institutions 
top_institutions <- applications_enriched %>%
  filter(!is.na(school)) %>%
  group_by(instellingserkenningscode, school) %>%
  summarise(
    applications = n(),
    conversions = sum(status == "ENROLLED", na.rm = TRUE),
    conversion_rate = mean(status == "ENROLLED", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(applications)) %>%
  head(10)

# Display top institutions
knitr::kable(top_institutions, 
             col.names = c("BRIN", "Institution", "Applications", 
                          "Enrollments", "Conversion Rate (%)"),
             digits = c(0, 0, 0, 0, 1))
```

```{r}
#| label: program-conversion-interactive

# Visualize program conversion rates (interactive)
plot_ly(top_programs, 
        x = ~reorder(Opleidingsnaam, conversion_rate), 
        y = ~conversion_rate, 
        type = "bar",
        text = ~paste0(round(conversion_rate, 1), "%"),
        hoverinfo = "text+x",
        marker = list(color = ~conversion_rate, 
                     colorscale = "Viridis")) %>%
  layout(title = "Conversion Rates by Top Programs",
         xaxis = list(title = ""),
         yaxis = list(title = "Conversion Rate (%)"),
         margin = list(b = 120),
         xaxis = list(tickangle = 45))
```

## 5. Geographic Analysis

If we have postcode data, we can analyze the geographic distribution of applications:

```{r}
#| label: geographic-analysis

# Count applications by postcode (first 2 digits)
postcode_data <- applications_enriched %>%
  filter(!is.na(postcodecijfers)) %>%
  mutate(postcode_prefix = substr(postcodecijfers, 1, 2)) %>%
  group_by(postcode_prefix) %>%
  summarise(
    applications = n(),
    conversions = sum(status == "ENROLLED", na.rm = TRUE),
    conversion_rate = mean(status == "ENROLLED", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(applications))

# Display top postcodes
head(postcode_data, 10) %>%
  knitr::kable(
    col.names = c("Postcode Region", "Applications", "Enrollments", "Conversion Rate (%)"),
    digits = c(0, 0, 0, 1)
  )
```

```{r}
#| label: postcode-visualization

# Sample visualization of applications by postcode region
ggplot(head(postcode_data, 20), 
       aes(x = reorder(postcode_prefix, applications), y = applications)) +
  geom_col(fill = "navy") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Applications by Postcode Region (Top 20)",
       x = "Postcode Region",
       y = "Number of Applications") +
  scale_y_continuous(labels = comma)
```

## 6. Crossing Patterns Analysis

Let's see the most common patterns of students applying to multiple institutions:

```{r}
#| label: crossing-patterns

# For students with multiple applications, analyze movement patterns
institution_patterns <- applications_enriched %>%
  filter(!is.na(school)) %>%
  group_by(bsnhash, schooljaar) %>%
  filter(n() > 1) %>%  # Only students with multiple applications
  summarise(
    institutions = paste(sort(unique(school)), collapse = " → "),
    is_enrolled = any(status == "ENROLLED"),
    .groups = "drop"
  ) %>%
  group_by(institutions) %>%
  summarise(
    student_count = n(),
    conversion_count = sum(is_enrolled),
    conversion_rate = mean(is_enrolled) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(student_count))

# Display top crossing patterns
head(institution_patterns, 10) %>%
  knitr::kable(
    col.names = c("Institution Pattern", "Students", "Enrollments", "Conversion Rate (%)"),
    digits = c(0, 0, 0, 1)
  )
```

## 7. Temporal Analysis

Let's examine application timing and its relationship to conversion:

```{r}
#| label: temporal-analysis

# Convert dates to proper format
applications_temporal <- applications_enriched %>%
  mutate(begindatum = as.Date(begindatum))

# Group by month
monthly_applications <- applications_temporal %>%
  mutate(month = floor_date(begindatum, "month")) %>%
  group_by(month) %>%
  summarise(
    applications = n(),
    conversions = sum(status == "ENROLLED", na.rm = TRUE),
    conversion_rate = mean(status == "ENROLLED", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  arrange(month)
```

```{r}
#| label: monthly-volume-plot

# Visualize monthly pattern
ggplot(monthly_applications, aes(x = month)) +
  geom_line(aes(y = applications), color = "blue", size = 1) +
  geom_point(aes(y = applications), color = "blue", size = 2) +
  theme_minimal() +
  labs(title = "Monthly Application Volumes",
       x = "Month",
       y = "Number of Applications") +
  scale_y_continuous(labels = comma)
```

```{r}
#| label: monthly-conversion-plot

# Visualize conversion rate over time
ggplot(monthly_applications, aes(x = month, y = conversion_rate)) +
  geom_line(color = "green4", size = 1) +
  geom_point(color = "green4", size = 2) +
  theme_minimal() +
  labs(title = "Monthly Conversion Rates",
       x = "Month",
       y = "Conversion Rate (%)") +
  scale_y_continuous(labels = function(x) paste0(x, "%"))
```

## Next Steps

This analysis forms the foundation for our Shiny app implementation. Key findings to highlight in the interactive dashboard:

1. Overall trends in application volume and conversion rates across years
2. The impact of multiple applications on conversion likelihood
3. Geographic patterns in application origin
4. Institution crossing patterns
5. Temporal patterns in application timing

The next steps are to:
1. Refine these analyses based on stakeholder feedback
2. Convert the static visualizations to interactive components
3. Implement the filtering functionality
4. Structure the app into logical modules following the Golem framework
